Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-09-24 arxiv/implications.tex
update
revise

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
We compare different choices of $M$ that belong to the following bounded set.
Let $\lambda_i$ be the $i$-th singular value of $M$.
Let $\mu_{\min} < \mu < \mu_{\max}$ be fixed values that do not grow with $p$.
{\begin{align*}
		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_i\le \mu_{\max}, \text{ for all } 1\le i\le p\right.},
\end{align*}}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.
\begin{proposition}[Covariate shift]\label{prop_covariate}
	For the setting of the example above, assume that $\beta_1 = \beta_2$, $\rho_1$ and $\rho_2$ are both greater than one.
	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
	We have that
		\[ g(\mu\id) \le \bigbrace{1+ \bigo{{\rho_2}/{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]
\end{proposition}
Proposition \ref{prop_covariate} shows that when the sample ratio is large, having no covariate shift gives the optimal performance for multi-task learning.
The proof of Proposition \ref{prop_covariate} can be found in Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

>>>>

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
We compare different choices of $M$ that belong to the following bounded set.
Let $\lambda_i$ be the $i$-th singular value of $M$.
Let $\mu_{\min} < \mu < \mu_{\max}$ be fixed values that do not grow with $p$.
{\begin{align*}
		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_i\le \mu_{\max}, \text{ for all } 1\le i\le p\right.},
\end{align*}}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.
\begin{proposition}[Covariate shift]\label{prop_covariate}
	For the setting of the example above, assume that $\beta_1 = \beta_2$, $\rho_1$ and $\rho_2$ are both greater than one.
	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
	We have that
		\[ g(\mu\id) \le \bigbrace{1+ \bigo{{\rho_2}/{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]
\end{proposition}
\cor

 \begin{claim}\label{claim_covar_shift}
		In the setting of Proposition \ref{prop_covariate}, for any $M\in \cal S_\mu$ we have that
		\[ g(M)=(1+\OO(p^{-\e}))\cdot \sigma^2  \bigtr{\Sigma_2(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }  \quad \text{w.h.p.} \]
	\end{claim}

%Finally, in this subsection we prove Proposition \ref{prop_covariate}, which shows that $\te(\hat{\beta}_2^{\MTL})$ is minimized approximately when $M$ is a scalar matrix, provided that there is enough source data.

\begin{proof}[Proof of Proposition \ref{prop_covariate}]

 Denote the minimizer of $g$ by 
$$M_0:=\argmin_{M\in \cal S_{\mu}}g(M).$$ 
We now calculate $g(M_0)$.

Now using Lemma \ref{lem_cov_shift}, we obtain that with high probability,
\begin{align}\label{gvar_extra}
g(M_0)= \frac{\sigma^2}{\rho_1+\rho_2}\cdot \frac1p\tr\left( \frac{1}{a_1(M_0)\cdot M_0^\top M_0 + a_2(M_0)}\right) \cdot \left(1 +\OO(p^{-\e})\right).
\end{align}
From equation \eqref{eq_a12extra}, it is easy to obtain the following estimates on $ a_1(M)$ and $a_2(M)$ for any $M\in \cal S_\mu$:
\be\label{est_a12extra}
\frac{\rho_1-1}{\rho_1+\rho_2} < a_1(M)<  \frac{\rho_1+\rho_2-1}{\rho_1+\rho_2},\quad a_2(M) < \frac{\rho_2}{\rho_1+\rho_2}.
\ee
Inserting \eqref{est_a12extra} into \eqref{gvar_extra} and using $ M_0^\top M_0\succeq \mu_{\min}^2$, we obtain that with high probability,
\begin{align}\label{approximateteM}
\left(1+\frac{\rho_2}{(\rho_1-1)\mu_{\min}^2}\right)^{-1}h(M_0) \cdot \left(1 - \OO(p^{-\e})\right) \le g(M_0) \le h(M_0) \cdot \left(1 +\OO(p^{-\e})\right),
\end{align}
where
$$h(M_0):=\frac{\sigma^2}{(\rho_1+\rho_2)a_1(M_0)}\cdot \frac1p\tr\left( \frac{1}{M_0^\top M_0}\right) .$$
%With these two bounds, we can easily conclude \eqref{approxteM}. 
%
%We have that the test error satisfies
%\be\label{approxteM}  te(M)\left(1 -  \frac{n_2}{n_1-p} \frac{1}{\lambda_p^2 + \frac{n_2}{n_1-p}}\right)  \le  \frac{\sigma^2}{n_1+n_2}\tr\left( \frac{1}{a_1M^\top M + a_2}\right) \le te(M),\ee
%where $\lambda_p$ is the smallest singular value of $p$ and
%$$te(M):= \frac{\sigma^2}{a_1(n_1+n_2)}\tr\left( \frac{1}{M^\top M}\right) .$$
%Moreover, for all $M$ satisfying \eqref{GMcons}, the minimum of $te(M)$ is attained when $M= a\id$.
By AM-GM inequality, we observe that 
$$\tr\left( \frac{1}{M^\top M}\right) = \sum_{i=1}^p\frac{1}{\lambda_i^2}$$
is minimized when $\lambda_1 = \cdots\lambda_p=\mu$ under the restriction $\prod_{i=1}^p\lambda_i\le \mu^p$. Hence we get that 
\be\label{AMGM} h(M_0) \le \frac{\sigma^2}{\mu^2 (\rho_1+\rho_2)a_1(M_0)}.\ee
On the other hand, when $M=\mu \id$, applying Lemma \ref{lem_cov_shift} we obtain that with high probability,
\begin{align}\label{gvar_extra2}
\begin{split}
g(\mu \id)&= \frac{\sigma^2}{\rho_1+\rho_2}\cdot \frac1p\tr\left( \frac{1}{\mu^2 a_1 (\mu\id) + a_2(\mu\id)}\right) \cdot \left(1 +\OO(p^{-\e})\right)\\
&\le \frac{\sigma^2}{\mu^2(\rho_1+\rho_2)a_1 (\mu\id)}.
\end{split}
\end{align}
Combining \eqref{est_a12extra}, \eqref{approximateteM}, \eqref{AMGM} and \eqref{gvar_extra2}, we conclude the proof.
%, we conclude that the sum $\sum_{i=1}^p\lambda_i^{-1}$ is smallest when $\lambda_1=\cdots=\lambda_p = a$.
\end{proof}
\nc

Proposition \ref{prop_covariate} shows that when the sample ratio is large, having no covariate shift gives the optimal performance for multi-task learning.
The proof of Proposition \ref{prop_covariate} can be found in Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

### 2020-09-24 arxiv/proof_examples.tex
wip
pf overview
revise

\section{Proof of the Special Case}\label{app_iso_cov}

@@

@@

@@

%--------old arguments
%With the same arguments as in Lemma \ref{lem_hat_v} we can show that \eqref{hatv_add1} holds, and under the choice of parameters \eqref{choiceofpara},
%\be \nonumber
%\begin{split}
%g(M)&=(1+\OO(n^{-\e}))\cdot d^2 \tr\left[(X_1^{\top}X_1)(X_1^{\top}X_1 +X_2^{\top}X_2)^{-1}\Sigma_2(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)\right] \\ 
%&+(1+\OO(n^{-\e}))\cdot \sigma^2  \bigtr{\Sigma_2(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }  =: \delta_{\bias}(M)+ g_{\vari}(M),
%\end{split}
%\ee
%with high probability for any constant $\e>0$. In fact, Lemma \ref{lem_hat_v} was proved assuming that $M=\id$, but its proof can be easily extended to the case with general $M\in \cal S_{\mu}$ by using that $\lambda(M)\in [\mu_{\min},\mu_{\max}]$. We omit the details here. 
%
%
%With Lemma \ref{lem_cov_shift}, we can obtain that with high probability,
%\begin{align}\label{gvar_extra}
%g_{\vari}(M_0)= \frac{\sigma^2}{\rho_1+\rho_2}\cdot \frac1p\tr\left( \frac{1}{a_1M_0^\top M_0 + a_2}\right)  +\OO(p^{-1/2+\e}),
%\end{align}
%where $(a_1,a_2)$ satisfies \eqref{eq_a12extra}. On the other hand, by Theorem \ref{thm_model_shift} we can estimate $\delta_{\bias}$ as
%\be \label{deltabeta_extra}
%\begin{split}
%&\left| \delta_{\bias}(M_0) - \frac{d^2\cdot \rho_1^2}{(\rho_1 + \rho_2)^2} \tr\left[\Sigma_1^{1/2}{M_0} \frac{(1 + a_3)\id + a_4 {M_0}^{\top}{M_0}}{(a_2 + a_1 {M_0}^{\top}{M_0})^2} {M_0}^{\top}\Sigma_1^{1/2}\right]\right| \le \delta + \OO(p^{1-\e}d^2),
%\end{split}
%\ee
%where $(a_3,a_4)$ satisfies \eqref{eq_a34extra} and $\delta$ is defined in \eqref{eq_deltaextra}.
%
%
%Next we derive some estimates for $a_i$, $i=1,2,3,4$. From equation \eqref{eq_a12extra}, it is easy to see that
%\be\label{est_a12extra}a_1\ge \frac{\rho_1-1}{\rho_1+\rho_2},\quad a_2\le \frac{\rho_2}{\rho_1+\rho_2}.\ee
%Then solving \eqref{eq_a34extra}, we get that
%\begin{align}
%0\le a_3&= \frac{ b_0 (\rho_1a_1^{-2}-b_2) + b_1^2 }{  \left(\rho_2  a_2^{-2}-  b_0\right)\left(\rho_1 a_1^{-2} -  b_2  \right) - b_1^2} \le \frac{ b_0 \rho_1a_1^{-2}  }{  \rho_1\rho_2  a_1^{-2}a_2^{-2}-  b_0\rho_1 a_1^{-2} - b_2\rho_2  a_2^{-2}}\nonumber\\
%&\le \frac{\rho_1}{\rho_1\rho_2-\rho_1 -\rho_2 },\label{est_a3extra}
%\end{align}
%and
%\begin{align}
%0\le a_4&= \frac{b_1\rho_2 a_2^{-2}}{\left(\rho_2  a_2^{-2}-  b_0\right)\left(\rho_1 a_1^{-2} -  b_2  \right) - b_1^2} \le  \frac{ b_1\rho_2 a_2^{-2} }{  \rho_1\rho_2  a_1^{-2}a_2^{-2}-  b_0\rho_1 a_1^{-2} - b_2\rho_2  a_2^{-2}} \nonumber\\
%&\le  \frac{ \rho_2 (b_1a_1^{2}) }{  \rho_1\rho_2 -  \rho_1  - \rho_2 } \le \frac{\rho_2\cdot \mu_{\min}^{-2}}{\rho_1\rho_2-\rho_1-\rho_2}=\OO\left( \frac{1}{\rho_1}\right),\label{est_a4extra}
%\end{align}
%where in the above derivations we also used that $b_1^2\le b_0 b_2$ by Cauchy-Schwarz inequality, $b_0 a_2^2 < 1$, $b_2 a_1^2 < 1$, and $b_1 a_1^2 \le \mu_{\min}^{-2}$.
%Inserting \eqref{est_a12extra} into \eqref{gvar_extra}, we obtain that with high probability,
%\begin{align}\label{approximateteM}
%\le g(M_0) \le  \frac{\sigma^2}{\rho_1+\rho_2}\cdot \frac1p\tr\left( \frac{1}{a_1M_0^\top M_0}\right) \cdot \left(1 +\OO(p^{-\e})\right)
%\end{align}
%where
%$$\wt g(M_0):=\frac{\sigma^2}{\rho_1+\rho_2}\cdot \frac1p\tr\left( \frac{1}{a_1M_0^\top M_0}\right)  +\frac{pd^2\cdot \rho_1^2}{(\rho_1 + \rho_2)^2} \cdot \frac1p\tr\left(\Sigma_1\frac{1 + a_3}{a_1^2 M_0^{\top}{M_0}} \right),$$
%and the error satisfies
%$$|\cal E|\le C\left( \frac{\rho_2}{\rho_1} +\rho_1^{-1/2} +\OO(p^{-\e})\right)\wt g(M_0).$$
%Here the constant $C>0$ depends only on $\mu_{\max}$, $\mu_{\min}$ and $\|\Sigma_1\|$, but otherwise does not depend on $\rho_1$ and $\rho_2$. 
%--------end old argument---------

\iffalse
In the simulation of Figure \ref{fig_covariate}, we observe the following two phases as  $n_1 / p$ increases.
When $n_1 \le n_2$, having complementary covariance matrices leads to lower test error compared to the case when $\Sigma_1 = \Sigma_2$.
When $n_1 > n_2$, having complementary covariance matrices leads to higher test error compared to the case when $\Sigma_1 = \Sigma_2$.
We provide a theoretical justification now.

@@

%\subsection{Proofs for Section \ref{sec_covariate}}\label{app_covariate}

\subsection{Denoising}

@@

\noindent{\bf Negative transfer.} On the other hand, if the following inequality holds,
\be\label{cond sigma12}
\begin{split}
&\sigma_2^2 \cdot \frac{\rho_1}{(\rho_2-1)(\rho_1 + \rho_2 -1)}\cdot \left(1 + \oo(1)\right) \\
&< pd^2 \cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}\al_-^2(\rho_1)  + (\sigma_1^2 -\sigma_2^2)\cdot \frac{\rho_1 (\rho_1+\rho_2)}{(\rho_1+\rho_2-1)^3} \al_-(\rho_1) ,
\end{split}
\ee
then we have $\delta_{\vari} < \delta_{\bias} + \delta_{\vari}^{(2)}$ w.h.p., which gives $\te(\hat{\beta}_t^{\MTL})>\te(\hat{\beta}_t^{\STL})$. We can solve \eqref{cond sigma12} to get
\begin{align*}
\sigma_1^2 > - pd^2 \cdot \rho_1 \al_-(\rho_1) +\sigma_2^2 \left[ 1+ \rho_1\Phi(\rho_1, \rho_2)\al_-^{-1}(\rho_1)\right]\cdot (1+\oo(1)).
\end{align*}
This proves the second claim of Proposition \ref{prop_var_transition}.
\end{proof}

\subsection{Data Efficiency}

>>>>

\section{Proof of the Special Case}\label{app_iso_cov}
\todo{switch the $a_3$ and $a_4$}

@@

\begin{proof}[Proof of Claim \ref{claim_covar_shift}]	
	With the same arguments as in Claim \ref{lem_hat_v}, we can show that \eqref{hatv_add1} holds. Moreover, if the parameters are chosen such that $p^{-1+c_0} \sigma^2 \le \kappa^2  \le p^{-\e_0-c_0}\sigma^2$ as in \eqref{choiceofpara}, we can simplify $g(M_0)$ as
\be \nonumber
\begin{split}
g(M_0)&=(1+\OO(p^{-\e}))\cdot \sigma^2  \bigtr{\Sigma_2(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }  \quad \text{w.h.p.}
\end{split}
\ee
 for some constant $\e>0$. In fact, Lemma \ref{lem_hat_v} was proved assuming that $M=\id$, but its proof can be easily extended to the case with general $M\in \cal S_{\mu}$ by using that $ \mu_{\min}\le \lambda_p(M)\le \lambda_1(M)\le \mu_{\max}$. We omit the details here. 
 \end{proof}

@@

\iffalse

@@

In the simulation of Figure \ref{fig_covariate}, we observe the following two phases as  $n_1 / p$ increases.
When $n_1 \le n_2$, having complementary covariance matrices leads to lower test error compared to the case when $\Sigma_1 = \Sigma_2$.
When $n_1 > n_2$, having complementary covariance matrices leads to higher test error compared to the case when $\Sigma_1 = \Sigma_2$.
We provide a theoretical justification now.

@@

\iffalse
\subsection{Label Denoising}

@@

\noindent{\bf Negative transfer.} On the other hand, if the following inequality holds,
\be\label{cond sigma12}
\begin{split}
&\sigma_2^2 \cdot \frac{\rho_1}{(\rho_2-1)(\rho_1 + \rho_2 -1)}\cdot \left(1 + \oo(1)\right) \\
&< pd^2 \cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}\al_-^2(\rho_1)  + (\sigma_1^2 -\sigma_2^2)\cdot \frac{\rho_1 (\rho_1+\rho_2)}{(\rho_1+\rho_2-1)^3} \al_-(\rho_1) ,
\end{split}
\ee
then we have $\delta_{\vari} < \delta_{\bias} + \delta_{\vari}^{(2)}$ w.h.p., which gives $\te(\hat{\beta}_t^{\MTL})>\te(\hat{\beta}_t^{\STL})$. We can solve \eqref{cond sigma12} to get
\begin{align*}
\sigma_1^2 > - pd^2 \cdot \rho_1 \al_-(\rho_1) +\sigma_2^2 \left[ 1+ \rho_1\Phi(\rho_1, \rho_2)\al_-^{-1}(\rho_1)\right]\cdot (1+\oo(1)).
\end{align*}
This proves the second claim of Proposition \ref{prop_var_transition}.
\end{proof}
\fi

\subsection{Labeling Efficiency}
